Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-06 paper/main.tex
pass through intro

@@

Available techniques split in two categories.  Tools
from the first one derive impressive bounds for
imperative programs, but they are not fully compositional.
Notable recent work like
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14} falls in this category.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
Second, while all imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no opportunity for sound interaction with the
tool, for example, manual proofs cannot be composed
with automatically inferred bounds.  One reason for
this impossibility to interact is that available tools
work on an abstract version of the source program. This
abstracted program has a lot more behaviors than the
original one and is usually expressed in a language that
does not feature memory operations.

In the second category, tools derive very tight
bounds on functional programs~\cite{Jost03,HoffmannAH12}.
They are highly compositional, scale for large
programs, and work directly on the syntax.  Despite all
these theoretical and practical advantages, it has been
a long-time open problem how to make them work in the
presence of negative integers, mutation and non-linear
control flow.  For example, the first challenge of
imperative programs is the absence of useful type system
to hook the analysis on.  Another issue is that these
tools associate a \emph{potential} directly to program
variables, doing this in imperative programs would
fail for loops as simple as \lstinline{for(i=x;i<y;i++)}
where $y - i$ decreases, but not $|i|$.
Mutation is yet another challenge, while functional
programs relate multiple fixed inputs (the typing
context) with one output (the expression typed),
imperative programs can change the value of variables
in scope, so the potential of these variables has to be
updated after mutations or, unsound bounds can be derived.

@@

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers
and differences of variables.  It is also the first
resource analysis tool for C that deals naturally with
recursive functions, resources that can be freed (e.g.
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:

@@

@@

>>>>

\newcommand{\jan}[1]{{\color{orange}[[{#1}]]}}

@@

Available techniques for deriving worst-case resource bounds
split in two categories.  Techniques
in the first category derive impressive bounds for
imperative programs, but they are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
In the second category, techniques derive tight whole-program
bounds on functional programs.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, despite all
these theoretical and practical advantages, it has been
a long-time open problem to make them work in the
presence of negative integers, mutation, and non-linear
control flow. As a result, they are not suitable for
typical imperative code.
%
\jan{I changed the order here a bit. Good style: First sentence is a
  summary of the following paragraph.}

Notable tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no possibility for sound user interaction with the
tool. For example, there is no concept of manual proofs for
resource bounds and no framework that could support composition
of manually derived bounds with automatically inferred bounds. 
One reason for this lack of interaction is that available tools
work on an abstract version of the source program. This
abstracted program has more behaviors than the
original one and is usually expressed in a language that
does not feature memory operations. 
%
\jan{@Quentin: What do you mean by ``more behaviors''?}

In the second category, are based on type systems for functional
programs and the potential method of amortized
analysis~\cite{Jost03,HoffmannAH12}.  The first challenge of
imperative programs is already the absence of a fine-grained type system to hook
the analysis on.  Another issue is that the techniques associate
\emph{potential} directly to program variables.  In
imperative programs, this would fail for loops as simple as
\lstinline{for(i=x;i<y;i++)} where $y - i$ decreases, but not $|i|$.
Mutation is yet another challenge.  While functional programs relate
multiple fixed inputs (the typing context) with one output (the
expression typed), imperative programs can change the value of
variables in scope, so the potential of these variables has to be
updated after mutations or, unsound bounds can be derived.
%
\jan{@Quentin: There is a mismatch between the previous two paragraphs.
  In the first one we talk about things that category one techniques
  can't do. In the second one we talk about ``challenges'' for the
  category two. We instead should just say what the state-of-the-art
  can't do. We also have to mention Hofmann/Jost ESOP'06 and Atkey ESOP'10.}

@@

Our automated analysis is able to infer linear
resource bounds on C programs with mutually-recursive functions and integer loops.
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions, resources that can become available during execution (e.g., when freeing
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:

@@

\jan{Here we need a description of the implementation and the
  experiments.  Message: You can immediately see the great benefits of
  the unification of the two categories in practice. It scales to
  larger programs than existing tools (for global bounds) and is as
  good on local bounds (for tricky loop and recursion patterns).}

@@

\jan{We also mention before that existing imperative tools don't allow
  for user interaction. We should pick this up again and say that we
  can do it. But then we have to mention the PLDI paper. My suggestion:
  \begin{itemize}
  \item mention PLDI'14 in the paragraph about the second category and
    say it has great flexibility for manual bound derivation but
    almost no support for automation
  \item just add another bullet to say that we have the same
    flexibility for manual bound derivation as in PLDI'14 and that
    manual and automatic bounds and be composed
  \end{itemize}
}
